---
title: "02_question2"
author: '44'
date: "2022-06-01"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
library(here)
library(nlme)
library(lme4)

library(tidyverse)
library(broom)
library(broom.mixed)
```


# Q2 Carry out the steps of the statistical analysis outlined below. 

Turn in **commented statistical code** used to obtain the results reported.
Use no more than **3 decimal places**.

Lower-limb amputation (LLA) results primarily from complications of 
severe peripheral artery disease, diabetes mellitus, or trauma. 

Sustained exercise has been suggested by previous studies as 
a promising rehabilitation target to improve the long-term health after LLA. 

Using telehealth, a clinical trial has been conducted at 
the Veterans Affairs Eastern Colorado Regional Amputation Center to 
test the potential of sustaining walking exercise using 
**exercise self-management (EXP)** versus an **attention-control education (CTL)** program. 

- The **primary outcome** is *step count per day* 
- A few other **secondary outcomes** are also of interest including *presence of any adverse events*, as binary outcome. 
- Outcomes were collected at **baseline (time=0)**, **6 months (time=6)** and **18 months (time=18)**. 
- The primary hypothesis of the trial was that **the intervention would improve daily steps in at least 1000 steps**. 
- Because randomization was **stratified by age ($< 60$ and $geq 60$)** and 
**level of amputation (below knee and above knee)**,
these two variables should be adjusted for in the analyses.
- The longitudinal dataset of the study is in the table `data-lla.csv` with data dictionary below:

  - treat: `treat = EXP` for individuals in EXP group, and `treat = CTL` for individuals in CTL group
  - stepc: Step count
  - amp.above: amp.above=1 for amputation above the knee, amp.above=0 for amputation below the knee
  - time: Month from baseline
  - older: `older=1` for individuals 60 or older, and `older=0` for individuals younger than 60 years of age
   
- Use R or SAS to answer the following questions regarding the hypothesis above.


## Part 2a

The primary hypothesis of the study was to test 
**whether there was a difference of more than 1000 steps in the change of step count/day at 6 months between exposure groups**. 

This particular type of trial is often referred to as a **superiority trial by a margin**. Carry out this hypothesis test and interpret your results for a clinical collaborator.

### i. 

Write the model formally in mathematical form. Allow for a difference in daily step count at baseline.

There are four different types of model can be used to solve only for 2a(i) - (iv): 

- a linear change score model (R code `fit_2_0`)

- a linear mixed model with random intercept (R code `fit_2_1`)

- a generalized linear mixed model for Poisson regression with random intercept (R code `fit_2_2`)

- a generalized estimate equation model for Poisson regression with compound symmetry correlation structure (not included due to only include marginal model)

However if we still need to consider the question for 2b, 
a linear mixed model or a generalized linear mixed model conditional on each individual are preferred.

There are pros and cons for each model:
For linear mixed model, the outcomes are counts as integers (not continuous), so there will be violations on the assumption for normally distributed residuals. 
however it is easier to calculate expected mean, standard error, and statistics for given groups, as well as, to test the primary hypothesis of the numeric difference between group.

For generalized linear mixed model on Poisson regressor, the model is perfect for count-outcomes. However the model can only provide the risk rate (log risk rate), it will be hard to calculate a meaningful standard error for given groups, so it is hard to test the primary hypothesis. Besides there might be over- or under-dispersion situation to mitigate the performance of the model.


According to figures and tables below in 2a(v):
There is nonlinear trend for the trajectories in treatment group,
it is reasonable to treat time as a categorical variable with (0, 6, and 18 three levels) other than a continuous variable.
Due to the parallel trajectory patterns for subjects from different groups, a random intercept will be applied for both models.
The time effects are different between different intervention groups, so a time, intervention interaction term must be included in the model.

I will also recommend to build regression model on risk difference in future studies, however the related materials are not included in the lectures.


#### Linear mixed model with only random intercept

$$
\begin{split}
Y_{ij} | b_{i} & \sim Normal(\mu_{ij}, \sigma_{\epsilon}^2)\\
\mu_{ij} & = X_{ij} \beta + Z_{i} b_{i}\\
Y_{ij} & = \mu_{ij} + \epsilon_{ij}\\
& = X_{ij} \beta + Z_{i} b_{i} + \epsilon_{ij}\\
& = \beta_0 + \beta_1 \times treat_i + \beta_2 \times I(time6)_{ij} + \\
& \ \ \ \ \ \beta_3 \times I(time18)_{ij} + \beta_4 \times  old_i + \beta_5 \times amp_i  + \\
&  \ \ \ \ \  \beta_6 \times [treat::I(time6)]_{ij} + \beta_7 \times [treat::I(time18)]_{ij} + Z_{i}b_{i} + \epsilon_{ij}\\
b_i & \sim Normal (0, \sigma_b^2) \\
\epsilon_{ij} & \sim Normal(0, \sigma_{\epsilon}^2)
\end{split}
$$

$$
\begin{split}
\pmb Y_i | \pmb b_i & \sim Normal(\pmb \mu_i, \pmb R_i)\\
\pmb \mu_i & = E[\pmb Y_i |\pmb b_i]  = \pmb X_i \pmb \beta + \pmb Z_i \pmb b_i\\
\pmb R_i & = Var[\pmb Y_i| \pmb b_i] = \sigma_{\epsilon}^2 \pmb I\\
\pmb \epsilon_i & \sim Normal (\pmb 0, \pmb R_i)\\
\pmb b_i & \sim Normal (\pmb 0, \pmb G_i)\\
\pmb V_i & = Cov[\pmb Y_i] = \pmb Z_i \pmb G_i \pmb Z_i^T + \pmb R_i\\
\pmb Y_i & \sim Normal(\pmb \mu_i, \pmb V_i)
\end{split}
$$

The $\pmb R_i$ and $\pmb G_i$ is identical for every individual subject $i$. Here we only use to distinguish the individual variance-covariance $\pmb R_i$ from a population level varaince-covariance $\pmb {\mathcal R}$ and $\pmb {\mathcal G}$

#### Poisson generalized linear mixed model

$$
\begin{split}
Y_{ij} | b_{i} & \sim Poisson(\lambda)\\
 \eta_{ij} & = log(E[Y_{ij}|b_{i}]) = log(\lambda | b_i) = X_{ij} \beta + Z_{i} b_{i}\\
& = \beta_0 + \beta_1 \times treat_i + \beta_2 \times I(time6)_{ij} + \\
& \ \ \ \ \ \beta_3 \times I(time18)_{ij} + \beta_4 \times  old_i + \beta_5 \times amp_i  + \\
&  \ \ \ \ \  \beta_6 \times [treat::I(time6)]_{ij} + \beta_7 \times [treat::I(time18)]_{ij} + Z_{i}b_{i}\\
b_i & \sim Normal (0, \sigma_b^2) \\
\end{split}
$$


$$
\begin{split}
\pmb Y_i | \pmb b_i & \sim Poisson(\lambda_i)\\
\lambda_i & = E[\pmb Y_i |\pmb b_i] \\
\pmb \eta_i & = \log (\lambda_i) = \log (E[\pmb Y_i |\pmb b_i]) = \pmb X_i \pmb \beta + \pmb Z_i \pmb b_i\\
\pmb b_i & \sim Normal (\pmb 0, \pmb G_i)\
\end{split}
$$


In such case, the interpretation for $\beta$s are risk ratios and the variance is not meaningful for the variance for numeric difference.
Hence we have to use a first order Taylor series approximation, similar to variance stablization.

$$
\begin{split}
log(Z) & \approx log(E[Z]) + (Z - E[Z]) E[Z]\\
E[log(Z)] & \approx logE[Z]\\
Var[log(Z)] & \approx E[Z]^{-2}Var[Z]\\
Var[Z] & = Var[log(Z)] \times e^{2 E[logZ]}\\
sd[Z] & = sd[log(Z)] \times e^{E[logZ]}
\end{split}
$$


### ii. 

Write down the null and alternative hypothesis of interest in terms of the model parameters specified in (a); keep the terms consistent with (a).


Null hypothesis: 

After adjusted for the other variables, there is no difference between LLA patients from the treatment group and control group, on *whether the change of step counts over 1000 at 6 month*.
Alternative hypothesis: 

After adjustment for baseline step counts and stratification on age and amputation location, treatment as a statistically significant effect on *whether the step counts over 1000 at 6 month* over control group.


For the older age group and amputation location above:

$H_0$ : $e^{\beta_0 + \beta_1 + \beta4 + \beta_5} (e^{\beta_2 + \beta_6} - 1) - e^{\beta_0 + \beta4 + \beta_5} (e^{\beta_2} - 1) \leq 1000$

$H_1$ : $e^{\beta_0 + \beta_1 + \beta4 + \beta_5} (e^{\beta_2 + \beta_6} - 1) - e^{\beta_0 + \beta4 + \beta_5} (e^{\beta_2} - 1) \geq 1000$


For the younger age group and amputation location above:

$H_0$ : $e^{\beta_0 + \beta_1 + \beta_5} (e^{\beta_2 + \beta_6} - 1) - e^{\beta_0 + \beta_5} (e^{\beta_2} - 1) \geq 1000$

$H_1$ : $e^{\beta_0 + \beta_1 + \beta_5} (e^{\beta_2 + \beta_6} - 1) - e^{\beta_0 + \beta_5} (e^{\beta_2} - 1) \leq 1000$


For the older age group and amputation location below:

$H_0$ : $e^{\beta_0 + \beta_1 + \beta_4} (e^{\beta_2 + \beta_6} - 1) - e^{\beta_0 + \beta_4} (e^{\beta_2} - 1) \geq 1000$

$H_1$ : $e^{\beta_0 + \beta_1 + \beta_4} (e^{\beta_2 + \beta_6} - 1) - e^{\beta_0 + \beta_4} (e^{\beta_2} - 1) \leq 1000$


For the younger age group and amputation location below:

$H_0$ : $e^{\beta_0 + \beta_1} (e^{\beta_2 + \beta_6} - 1) - e^{\beta_0} (e^{\beta_2} - 1) \geq 1000$

$H_1$ : $e^{\beta_0 + \beta_1} (e^{\beta_2 + \beta_6} - 1) - e^{\beta_0} (e^{\beta_2} - 1) \leq 1000$



### iii. Calculate the change (from baseline) in expected means and standard error of the step counts at 6 months for each the intervention and the control group.


```{r}
## import dataset data_lla
data_lla <- read_csv("data/data-lla.csv", 
                     col_types = cols(
                       ## amputation location as binary variable
                       amp.above = col_factor(levels = c("0", "1")),
                       ## older age or not as binary variable 
                       older = col_factor(levels = c("0", "1")))) 


data_lla_diff6 <- data_lla %>% 
  ## change the from long format into wide format
  pivot_wider(names_from = "time",
              values_from = "stepc") %>%
  ## calculated the different from time0 to time6
  mutate(step_diff = abs(`6` - `0`)) %>% 
  ## binary outcome of whether the change of step is over 1000
  mutate(step_diff_1k = case_when(step_diff >= 1000 ~ 1,
                                  step_diff < 1000 ~ 0),
         baseline = `0`)

# View(data_lla_diff6)
```


#### using change score model

```{r}
fit_2_0 <- lm(step_diff ~ 1 + treat + older + amp.above,
               # control = fit_control,
               data = data_lla_diff6)

## to get the coefficients
beta_2_0 <- coef(fit_2_0) %>% round(digits = 3)
beta_2_0
## to get the variance-covariance
vcov_2_0 <- vcov(fit_2_0) %>% round(digits = 3)
vcov_2_0


##                  [1]       [2]       [3]       [4]       
## coefs           (Int)      treatEXP  old1      amp.above1
con_trt_old_abv <- c(1,        1,        1,        1)
con_trt_old_blw <- c(1,        1,        1,        0)
con_trt_yng_abv <- c(1,        1,        0,        1)
con_trt_yng_blw <- c(1,        1,        0,        0)

con_ctl_old_abv <- c(1,        0,        1,        1)
con_ctl_old_blw <- c(1,        0,        1,        0)
con_ctl_yng_abv <- c(1,        0,        0,        1)
con_ctl_yng_blw <- c(1,        0,        0,        0)

## this is the contrast of all the 
contr1 <- cbind(con_trt_old_abv,
                con_trt_old_blw,
                con_trt_yng_abv,
                con_trt_yng_blw,
                con_ctl_old_abv,
                con_ctl_old_blw,
                con_ctl_yng_abv,
                con_ctl_yng_blw)

contr1_1 <- cbind(con_trt_old_abv - con_ctl_old_abv,
                  con_trt_old_blw - con_ctl_old_blw,
                  con_trt_yng_abv - con_ctl_yng_abv,
                  con_trt_yng_blw - con_ctl_yng_blw)

rownames(contr1) <- rownames(vcov_2_0)
```


```{r}
mu_2_0 <- t(contr1) %*% beta_2_0 %>%
  as.data.frame() %>%
  dplyr::select("expected_mean" = 1)
## contrast variance covariance matrix
sd_2_0 <- t(contr1) %*% vcov_2_0 %*% contr1 %>%
  diag() %>% sqrt()
cbind(mu_2_0, "standard_error" = sd_2_0) %>%
  round(3)

mu_2_01 <- t(contr1_1) %*% beta_2_0 %>%
  as.data.frame() %>%
  dplyr::select("expected_mean" = 1)
## contrast variance covariance matrix
sd_2_01 <- t(contr1_1) %*% vcov_2_0 %*% contr1_1 %>%
  diag() %>% sqrt()
cbind(mu_2_01, "standard_error" = sd_2_01) %>%
  round(3)

qnorm(0.05, 1081.156,	0.811)	
```



#### using linear mixed model

```{r}
data_lla_tf <- data_lla %>%
  mutate(time = as.factor(time))


fit_2_1 <- lmer(stepc ~ 1 + treat * time + older + amp.above + 
                   + (1|id),
                 # control = fit_control,
                 data = data_lla_tf)

## to get the coefficients
beta_2_1 <- (summary(fit_2_1)$coefficients[, 1]) %>% round(digits = 3)
# beta_2_1
## to get the variance-covariance
vcov_2_1 <- vcov(fit_2_1) 
# vcov_2_1

## lmm contrast---------------------------------------------------------------------------
##                      [0]    [1]       [2]   [3]     [4]     [5]       [6]           [7]
## coefs               (Int)  treatEXP  time6 time18  older1  amp.abv1  trtEXP:time6  trtEXP:time18
con_trt_t0 <-   c(1,    1,        0,    0,      1,      1,        0,            0)
con_trt_t6 <-   c(1,    1,        1,    0,      1,      1,        1 * 1,        0)
con_trt_t18 <-  c(1,    1,        0,    1,      1,      1,        0,            1 * 1)
con_ctl_t0 <-   c(1,    0,        0,    0,      1,      1,        0,            0)
con_ctl_t6 <-   c(1,    0,        1,    0,      1,      1,        0,            0)
con_ctl_t18 <-  c(1,    0,        0,    1,      1,      1,        0,            0)
con_trt_t6_t0 <- con_trt_t6 - con_trt_t0
con_ctl_t6_t0 <- con_ctl_t6 - con_ctl_t0
con_trt_t6_t0 - con_ctl_t6_t0
```

```{r}
contrast0 <- cbind(con_trt_t6_t0, con_ctl_t6_t0)
mu_lm <- t(contrast0) %*% beta_2_1 
## standard error for each group
sd_lm <- t(contrast0) %*% vcov_2_1 %*% contrast0 %>%
  diag() %>% sqrt()

cbind(mu_lm, sd_lm) %>%
  as.data.frame() %>%
  dplyr::select("expected mean" = 1,
                "standard_error" = 2) %>%
  round(3)


t <- (1140.121 - (-58.962))/ 1.062; t
```

```{r}
con_trt_t6_t0 <- con_trt_t6 - con_trt_t0
con_ctl_t6_t0 <- con_ctl_t6 - con_ctl_t0

con_tvc_t6_t0 <- con_trt_t6_t0 - con_ctl_t6_t0
mu_2_3 <- t(con_tvc_t6_t0 ) %*% beta_2_1 %>%
  as.data.frame() %>%
  dplyr::select("expected_mean" = 1)
## contrast variance covariance matrix
sd_2_3 <- t(con_tvc_t6_t0 ) %*% vcov_2_1 %*% con_tvc_t6_t0  %>%
  diag() %>% sqrt()

cbind(mu_2_3, "standard_error" = sd_2_3) %>%
  round(3)

qnorm(0.05, mean = 1199.083, sd = 1.501)
pnorm(1000, mean = 1199.083, sd = 1.501)
```

The 95% confidence interval for *the change of step counts over 1000 at 6 month* is $(1196.614, \infty)$

#### using generalized linear mixed model

```{r}
fit_2_2 <- glmer(stepc ~ 1 + treat * time + older + amp.above + 
                   + (1|id),
                 family = "poisson",
                 # control = fit_control,
                 data = data_lla_tf)

## to get the coefficients
beta_2_2 <- (summary(fit_2_2)$coefficients[, 1]) %>% round(digits = 3)
# beta_2_2
## to get the variance-covariance
vcov_2_2 <- vcov(fit_2_2) 
# vcov_2_2
```



```{r}
## old amputation above ---------------------------------------------------------------------------
##                      [0]    [1]       [2]   [3]     [4]     [5]       [6]           [7]
## coefs               (Int)  treatEXP  time6 time18  older1  amp.abv1  trtEXP:time6  trtEXP:time18
con_old_abv_trt_t0 <-   c(1,    1,        0,    0,      1,      1,        0,            0)
con_old_abv_trt_t6 <-   c(1,    1,        1,    0,      1,      1,        1 * 1,        0)
con_old_abv_trt_t18 <-  c(1,    1,        0,    1,      1,      1,        0,            1 * 1)
(con_old_abv_trt_t6_t0 <- con_old_abv_trt_t6 - con_old_abv_trt_t0)
(con_old_abv_trt_t18_t0 <- con_old_abv_trt_t18 - con_old_abv_trt_t0)
con_old_abv_ctl_t0 <-   c(1,    0,        0,    0,      1,      1,        0,            0)
con_old_abv_ctl_t6 <-   c(1,    0,        1,    0,      1,      1,        0,            0)
con_old_abv_ctl_t18 <-  c(1,    0,        0,    1,      1,      1,        0,            0)
(con_old_abv_ctl_t6_t0 <- con_old_abv_ctl_t6 - con_old_abv_ctl_t0)
(con_old_abv_ctl_t18_t0 <- con_old_abv_ctl_t18 - con_old_abv_ctl_t0)
## young amputation above ---------------------------------------------------------------------------
##                      [0]    [1]       [2]   [3]     [4]     [5]       [6]           [7]
## coefs               (Int)  treatEXP  time6 time18  older1  amp.abv1  trtEXP:time6  trtEXP:time18
con_yng_abv_trt_t0 <-   c(1,    1,        0,    0,      1,      0,        0,            0)
con_yng_abv_trt_t6 <-   c(1,    1,        1,    0,      1,      0,        1 * 1,        0)
con_yng_abv_trt_t18 <-  c(1,    1,        0,    1,      1,      0,        0,            1 * 1)
con_yng_abv_trt_t6_t0 <- con_yng_abv_trt_t6 - con_yng_abv_trt_t0
con_yng_abv_trt_t18_t0 <- con_yng_abv_trt_t18 - con_yng_abv_trt_t0
con_yng_abv_ctl_t0 <-   c(1,    0,        0,    0,      1,      0,        0,            0)
con_yng_abv_ctl_t6 <-   c(1,    0,        1,    0,      1,      0,        0,            0)
con_yng_abv_ctl_t18 <-  c(1,    0,        0,    1,      1,      0,        0,            0)
con_yng_abv_ctl_t6_t0 <- con_yng_abv_ctl_t6 - con_yng_abv_ctl_t0
con_yng_abv_ctl_t18_t0 <- con_yng_abv_ctl_t18 - con_yng_abv_ctl_t0
## old amputation below ---------------------------------------------------------------------------
##                      [0]    [1]       [2]   [3]     [4]     [5]       [6]           [7]
## coefs               (Int)  treatEXP  time6 time18  older1  amp.abv1  trtEXP:time6  trtEXP:time18
con_old_blw_trt_t0 <-   c(1,    1,        0,    0,      0,      1,        0,            0)
con_old_blw_trt_t6 <-   c(1,    1,        1,    0,      0,      1,        1 * 1,        0)
con_old_blw_trt_t18 <-  c(1,    1,        0,    1,      0,      1,        0,            1 * 1)
con_old_blw_trt_t6_t0 <- con_old_blw_trt_t6 - con_old_blw_trt_t0
con_old_blw_trt_t18_t0 <- con_old_blw_trt_t18 - con_old_blw_trt_t0
con_old_blw_ctl_t0 <-   c(1,    0,        0,    0,      0,      1,        0,            0)
con_old_blw_ctl_t6 <-   c(1,    0,        1,    0,      0,      1,        0,            0)
con_old_blw_ctl_t18 <-  c(1,    0,        0,    1,      0,      1,        0,            0)
con_old_blw_ctl_t6_t0 <- con_old_blw_ctl_t6 - con_old_blw_ctl_t0
con_old_blw_ctl_t18_t0 <- con_old_blw_ctl_t18 - con_old_blw_ctl_t0
## young amputation below ---------------------------------------------------------------------------
##                      [0]    [1]       [2]   [3]     [4]     [5]       [6]           [7]
## coefs               (Int)  treatEXP  time6 time18  ynger1  amp.abv1  trtEXP:time6  trtEXP:time18
con_yng_blw_trt_t0 <-   c(1,    1,        0,    0,      0,      0,        0,            0)
con_yng_blw_trt_t6 <-   c(1,    1,        1,    0,      0,      0,        1 * 1,        0)
con_yng_blw_trt_t18 <-  c(1,    1,        0,    1,      0,      0,        0,            1 * 1)
con_yng_blw_trt_t18_t0 <- con_yng_blw_trt_t18 - con_yng_blw_trt_t0
con_yng_blw_trt_t6_t0 <- con_yng_blw_trt_t6 - con_yng_blw_trt_t0
con_yng_blw_ctl_t0 <-   c(1,    0,        0,    0,      0,      0,        0,            0)
con_yng_blw_ctl_t6 <-   c(1,    0,        1,    0,      0,      0,        0,            0)
con_yng_blw_ctl_t18 <-  c(1,    0,        0,    1,      0,      0,        0,            0)
con_yng_blw_ctl_t18_t0 <- con_yng_blw_ctl_t18 - con_yng_blw_ctl_t0
con_yng_blw_ctl_t6_t0 <- con_yng_blw_ctl_t6 - con_yng_blw_ctl_t0

# contast_old_abv_t06 <- cbind(con_old_abv_trt_t6_t0, con_old_abv_ctl_t6_t0)
# contast_yng_abv_t06 <- cbind(con_yng_abv_trt_t6_t0, con_yng_abv_ctl_t6_t0)
# contast_old_blw_t06 <- cbind(con_old_blw_trt_t6_t0, con_old_blw_ctl_t6_t0)
# contast_yng_blw_t06 <- cbind(con_yng_blw_trt_t6_t0, con_yng_blw_ctl_t6_t0)

contrast1 <- cbind(con_old_abv_trt_t6_t0,
                   con_yng_abv_trt_t6_t0,
                   con_old_blw_trt_t6_t0,
                   con_yng_blw_trt_t6_t0,
                   con_old_abv_ctl_t6_t0,
                   con_yng_abv_ctl_t6_t0,
                   con_old_blw_ctl_t6_t0,
                   con_yng_blw_ctl_t6_t0)
rownames(contrast1) <- rownames(vcov_2_2)
```

```{r}
con_ctl_t0 <- cbind(con_old_abv_trt_t0,
                    con_yng_abv_trt_t0,
                    con_old_blw_trt_t0,
                    con_yng_blw_trt_t0,
                    con_old_abv_ctl_t0,
                    con_yng_abv_ctl_t0,
                    con_old_blw_ctl_t0,
                    con_yng_blw_ctl_t0)

con_ctl_t6 <- cbind(con_old_abv_trt_t6,
                    con_yng_abv_trt_t6,
                    con_old_blw_trt_t6,
                    con_yng_blw_trt_t6,
                    con_old_abv_trt_t6,
                    con_yng_abv_trt_t6,
                    con_old_blw_trt_t6,
                    con_yng_blw_trt_t6)

mu_ctl_t0 <- t(con_ctl_t0) %*% beta_2_2 %>%
  exp() %>%
  as.data.frame() %>%
  dplyr::select("t0" = 1)
sd_ctl_t0 <- t(con_ctl_t0) %*% vcov_2_2 %*% con_ctl_t0 %>%
  diag() %>% sqrt()
mu_ctl_t6 <- t(con_ctl_t6) %*% beta_2_2 %>%
  exp() %>%
  as.data.frame() %>%
  dplyr::select("t6" = 1)
sd_ctl_t6 <- t(con_ctl_t6) %*% vcov_2_2 %*% con_ctl_t6 %>%
  diag() %>% sqrt()


mu_2_2 <- t(contrast1) %*% beta_2_2 %>%
  as.data.frame() %>%
  dplyr::select("log_rate" = 1) %>%
  mutate(rate = exp(log_rate)) 
## contrast variance covariance matrix
sd_2_2 <- t(contrast1) %*% vcov_2_2 %*% contrast1 %>%
  diag() %>% sqrt()

result2 <- cbind(mu_2_2,
      sd_log_rate = sd_2_2,
      mu_ctl_t0,
      mu_ctl_t6) %>% 
  mutate(diff_t6_t0 = (rate - 1) * t0,
         sd_t6_t0 = abs(rate * sd_2_2 * t0)) %>%
  round(3) 
result2
```

According to the model, the effect size for older age and amputation location is very small, hence the expected mean values and the standard errors of *the change of step counts over 1000 at 6 month from baseline* are very close to each others among different age group and amputation location groups. 

As we can see the difference on linear mixed model and generalized linear mixed model, the difference between t0 to t6 in linear mixed model is larger than generalized linear mixed model. There might be a over-dispersion for generalized linear mixed model. 

### iv.

Calculate the difference (EXP - CLT) in change of expected means in the two intervention groups. Report a 1-sided 95% confidence interval for this difference.


```{r}
result3a <- result2[1:4, ]
result3b <- result2[5:8, ]

result4 <- data.frame(diff_trt_ctl = result3a$diff_t6_t0 - result3b$diff_t6_t0,
                      sd_trt_ctl = sqrt(result3a$sd_t6_t0^2 + result3b$sd_t6_t0^2)) %>%
  mutate(q05 = qnorm(0.05, mean = diff_trt_ctl, sd = sd_trt_ctl)) %>%
  round(3)

result4
```



### v. 

Make a graph, and include any appropriate descriptive statistics, 
that you would present as evidence related to the primary hypothesis of interest.

```{r}
plot_lla <- data_lla %>%
  mutate(older = case_when(older == 0 ~ "Young",
                           older == 1 ~ "Old"),
         amp.above = case_when(amp.above == 1 ~ "Above",
                               amp.above == 0 ~ "Below")) %>%
  ggplot(aes(x = time, 
             y = stepc,
             group = id,
             color = treat)) +
  geom_line() +
  facet_wrap(older ~ amp.above)

plot_lla_log <- data_lla %>%
  mutate(older = case_when(older == 0 ~ "Young",
                           older == 1 ~ "Old"),
         amp.above = case_when(amp.above == 1 ~ "Above",
                               amp.above == 0 ~ "Below")) %>%
  ggplot(aes(x = time, 
             y = log(stepc),
             group = id,
             color = treat)) +
  geom_line() +
  facet_wrap(older ~ amp.above)

plot_lla + theme_bw()
plot_lla_log + theme_bw()
```


```{r}
plot_lla2 <- data_lla %>%
  mutate(older = case_when(older == 0 ~ "Young",
                           older == 1 ~ "Old"),
         amp.above = case_when(amp.above == 1 ~ "Above",
                               amp.above == 0 ~ "Below")) %>% 
  pivot_wider(names_from = time, 
              names_prefix = "t", 
              values_from = stepc) %>%
  mutate(diff = t6 - t0) %>% 
  ggplot(aes(y = diff, 
             x = treat,
             color = treat)) +
  geom_boxplot() +
  facet_wrap(older ~ amp.above)

plot_lla2 + theme_bw()
```



### vi. 

Write your conclusion of the hypothesis test in (b).

Based on the linear mixed model `fit_2_1` as well as the generalized linear mixed model `fit_2_2`, conditional on given individual adjusted for age group and operational amputation location, we can reject the null hypothesis and claim there is a very highly statistically significant difference on *the change of step counts over 1000 at 6 month* between two intervention groups (p-value < 0.05). 

### vii.

Comment on your assumptions of the correlation structure for the repeated measurements within subject.

The correlation structure is assumed to be compound symmetric structure, with ICC as $\rho = \frac {\sigma_b^2} {\sigma_b^2 + \sigma_{\epsilon}^2}$.


### viii. 

Write a summary paragraph interpreting the results for a clinical collaborator.

For linear mixed model, on average in population level, after adjusted for age group and amputation location, we found that *the change of step counts over 1000 at 6 month* in treatment group is 1199 steps more than the control group (sd = 1.5). Hence *the change of step counts over 1000 at 6 month* from different intervention groups is larger than 1000 with both practical and statistical significance. Therefore we can reject the null hypothesis and claim that after adjustment for baseline step counts and stratification on age and amputation location, treatment as a statistically significant effect on *whether the step counts over 1000 at 6 month* over control group.

For generalized linear mixed model with Poisson regression, the conclusion is very similar to linear mixed model.


## Part 2b

A secondary hypothesis of interest was whether or not the change in step counts was sustained from 6 to 18 months, specifically, interest was in testing the difference between **M18-M0** and **M6-M0** in the intervention group. 

Carry out the appropriate hypothesis to test whether there was a sustained change and interpret the results for a clinical collaborator.

For linear mixed model:

$H_0$: There is no difference between *the change of step counts over 1000 at 6 month from baseline* and  *the change of step counts over 1000 at 18 month from baseline* 

equation form: $\beta_3 + \beta_7 - \beta_2 - \beta_6 = 0$

$H_1$: There is a statistically significant difference between *the change of step counts over 1000 at 6 month from baseline* and  *the change of step counts over 1000 at 18 month from baseline*

equation form: $\beta_3 + \beta_7 - \beta_2 - \beta_6 neq 0$

```{r}
## lmm contrast---------------------------------------------------------------------------
##                      [0]    [1]       [2]   [3]     [4]     [5]       [6]           [7]
## coefs               (Int)  treatEXP  time6 time18  older1  amp.abv1  trtEXP:time6  trtEXP:time18
con_trt_t0 <-   c(1,    1,        0,    0,      1,      1,        0,            0)
con_trt_t6 <-   c(1,    1,        1,    0,      1,      1,        1 * 1,        0)
con_trt_t18 <-  c(1,    1,        0,    1,      1,      1,        0,            1 * 1)

con_trt_t6_t0 <- con_trt_t6 - con_trt_t0
con_trt_t18_t0 <- con_trt_t18 - con_trt_t0
contrast4 <- con_trt_t18_t0 - con_trt_t6_t0
contrast4

mu_b <- t(contrast4) %*% beta_2_1 
mu_b
## contrast variance covariance matrix
cov_b <- t(contrast4) %*% vcov_2_1 %*% contrast4 %>% as.matrix()


## with both point estimates and standard deviation 
## an anova or pairwise comparison can be performed
W0 <- mu_b^2 * solve(cov_b)
pchisq(W0, df = 1, lower.tail = FALSE)
```

For generalized linear mixed model:

$H_0$: There is no difference between *the change of step counts over 1000 at 6 month from baseline* and  *the change of step counts over 1000 at 18 month from baseline* 

equation form: $e^{\beta_3 + \beta_7 - \beta_2 - \beta_6} = 1$

$H_1$: There is a statistically significant difference between *the change of step counts over 1000 at 6 month from baseline* and  *the change of step counts over 1000 at 18 month from baseline*

equation form: $e^{\beta_3 + \beta_7 - \beta_2 - \beta_6} \neq 1$

```{r}
contrast5 <- con_old_abv_trt_t18_t0 - con_old_abv_trt_t6_t0
contrast5

mu_b2 <- t(contrast5) %*% beta_2_2 
exp(mu_b2)
## contrast variance covariance matrix
cov_b2 <- t(contrast5) %*% vcov_2_2 %*% contrast5 %>% as.numeric()

W1 <- mu_b2^2 / cov_b2
pchisq(W1, df = 1, lower.tail = FALSE)
```

As we can see in both linear mixed model and generalized linear mixed model, with Wald tests, the Wald statistics follow $\chi^2$ distribution. There is a change of `r mu_b` steps change between *the change of step counts over 1000 at 6 month from baseline* and  *the change of step counts over 1000 at 18 month from baseline* in linear mixed model with (p-value < 0.001); *the change of step counts over 1000 at 18 month from baseline* is `r (1 - exp(mu_b2)) * 100`% less than the *the change of step counts over 1000 at 6 month from baseline* in Poisson generalized linear mixed model with (p-value < 0.001). In both cases, the difference is statistically significant, hence we can reject the null hypothesis and claim that *the change in step counts* was not sustained from 6 to 18 months.
