---
title: "03_question3"
author: '44'
date: "2022-06-02"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q3

Assume that $(Y_i, X_i)$, $i = 1, ... , n$ are independent observations, where $Y_i$ is binary and $X_i$ is a $d$−dimensional vector of covariates.

We wish to fit a logistic regression model of $Y$ to $X$, i.e., $P(Y_i = 1) = \frac {exp(X_i \beta)} {(1 + exp(X_i \beta))}$ where $\beta$ are the regression coefficients.

Note that we are fitting a model without an intercept

### (a). Determine a generative model based on the multivariate normal distribution 

for $X|Y$ with $d = 3$ and $n = 102$ that leads to estimates of $\beta$ that are either negative infinity or infinity. We will refer to this as a nonconvergence situation.

List out the following values: 
the mean vector and variance-covariance matrix of $X|Y$ and $p$, the probability of $Y = 1$.


$$
\begin{split}
Y_i | X_i \sim Bernoulli(p_i)\\
logit(E[Y_i]) = logit(p_i) = X_i \beta\\
p_i(X_i) = \frac {\exp (X_i\beta) } {1 + \exp (X_i \beta)}
\end{split}
$$

$$
\begin{split}
f(x, y| \beta) & = L(\beta; X, Y) \\
& =  \prod_{i = 1}^n  \Big(p_i(x_i)\Big)^{y_i}\Big(1- p_i(x_i)\Big)^{1 - y_i}\\
& =  \prod_{i = 1}^n  \Bigg(\frac {\exp (X_i\beta) } {1 + \exp (X_i \beta)}\Bigg)^{y_i}\Bigg(1- \frac {\exp (X_i\beta) } {1 + \exp (X_i \beta)}\Bigg)^{1 - y_i}\\
log L(\beta; X, Y)  & = \sum_{i=1}^nlog\Bigg(1- \frac {\exp (X_i\beta) } {1 + \exp (X_i \beta)}\Bigg) + \sum_{i=1}^n y_i X_i\beta\\
\frac {\nabla log L(\beta; X, Y)} {\nabla \beta} & = \sum_{i=1}^n\Bigg(y_i - \frac {\exp (X_i\beta) } {1 + \exp (X_i \beta)}\Bigg) X_i \stackrel {set} {=} 0\\
\forall i, & \ y_i = \frac {\exp (X_i \hat \beta) } {1 + \exp (X_i \hat \beta)} 
\end{split}
$$

In a general situation, we will not be able to solve exactly to get a closed form solution, because binary outcome $y_i$ can not be estimate by $p_i(X_i) = \frac {\exp (X_i\beta) } {1 + \exp (X_i \beta)}$. For most of the time, numerical methods must applied. However, in some rare cases ($\beta$ is noncovergence, as $-\infty$ or $\infty$), $p_i(X_i) = \frac {\exp (X_i\beta) } {1 + \exp (X_i \beta)}$ can take the exact value of $y_i$ as either $0$ or $1$. This means the value of $Y$ can separate a given $X$ completely. Then we will have a situation of complete separation. In such case, the more extreme the value of $\beta$ takes, the larger the likelihood is. Hence the maximum likelihood estimate on the parameter $\beta$ does not exist, but take the value of $-\infty$ or $\infty$.


the mean vector and variance-covariance matrix of $X|Y$ and $p$, the probability of $Y = 1$.

I will set $x1$ is correlated with $y$ with mean $E[x_1] = 0$, and make sure $\pmb X \pmb \beta = 0$. 
In such case, we have $logit(p) = \pmb X \pmb \beta = 0$. So we set $\mu_1 = E[X_1] = 0$, and $E[x_2 * \beta_2 + x_3 \times \beta_3] = 0$ to generate $Y_i \sim Bernoulli(0.5)$. Then we set $x_1$ perfectly complete separated by $Y_i$, such that when $P(X_1 > 0 | Y = 1) = 1$.

$$
\begin{split}
& p = E[Y] = 0.5\\
& logit(p) = \pmb X \pmb \beta = 0\\
& \pmb \beta = (1, 0.5, -0.25)\\
& \pmb X \sim MVN(\pmb \mu, \pmb \Sigma)\\
& \pmb \mu = \begin{pmatrix}
  \mu_1 = 0 \\ \mu_2 = 1 \\ \mu_3 = 2
\end{pmatrix}, \
\pmb \Sigma = \begin{pmatrix}
  \sigma_1^2 = 0.25 & 0 & 0\\
  0 & \sigma_2^2  = 0.25 & 0\\
  0 & 0 & \sigma_3^2 = 0.25
\end{pmatrix}
\end{split}
$$



### (b). Give a 2-3 sentence description which describes when we can expect nonconvergence.
The outcome variable separates certain predictor variables completely and almost perfectly. 

There is no pair $(Y, X)$ with give $Y$ value and $X$ belongs to certain region, such that the frequency is zero in the sample, for example $P(Y = 1| X_{sub} \in (-\infty, 0)) = 0$ and $P(Y = 1| X_{sub} \in (0, \infty)) = 1$. 

In such situation, maximum likelihood estimate on the parameter $\beta$ does not exist.

### (c). Give a 2-3 sentence description which describes how nonconvergence relates to multicollinearity.

In both situations, the model does not have reasonable estimates for predictor variables, but remains valid prediction for the outcome.

The noncovergence situation is a special type of collinearity, after link function transformation on $y$, there is a linear relationship with $X_2, X_3$, $logit(P(Y)) = x_2 \beta_2 + x_3 \beta_3$. Meanwhile, we know $Y = I(X_1\in S)$ for separation case. Hence $logit(P(X_1 \in S)) = x_2 \beta_2 + x_3 \beta_3$ is collinearized with $X$ belongs to certain set. 

Although in the noncovergence cases, the outcome $y$ is not collinear with $X_{sub}$, a subset of $X$, but a binary variable of $I(X_{sub} \in S)$.

### (d). Modify the model in (a) to achieve convergence. 

Based on the modified model, fit a logistic regression of $y$ on $X$ using 100000 observations from $y = 1$ and 100000 observations from $y = 0$. 

Report the estimated regression coefficients and standard errors.

Assuming $E[Y] = \hat p = 0.5$, such that $logit(\hat p) = X\beta = 0$, we can discard the predictor $x_1$, which is correlated with outcomes $Y$. If we only use the other two predictors $x_2, x_3$, the model should converge.

```{r "sim_method1", eval=FALSE, include=TRUE}
library(tidyverse)
set.seed(55555)

n <- 100000
N <- n * 10
beta1 <- 1
beta2 <- 0.5
beta3 <- -0.25

x1abs <- c(abs(rnorm(n, mean = 0, sd = 0.5)))
x2 <- c(rnorm(2 * N, mean = 1, sd = 0.5))
x3 <- c(rnorm(2 * N, mean = 2, sd = 0.5))

beta23 <- as.matrix(c(beta2, beta3))
X23 <- as.matrix(cbind(x2, x3))
# str(beta23); dim(beta23)
# str(X23); dim(X23)

logit_p <- X23 %*% beta23
mean(logit_p) ## [1] 4.258361e-05
p_hat <- 1 / (1 + exp(-logit_p))
mean(p_hat) ## [1] 0.500011

y <- map(p_hat, 
         ~rbinom(n = 1, size = 1, prob = .)) %>% 
  unlist()

data_3_0 <- data.frame(y, x2, x3) %>%
  ## simulate x1 which can be separate by outcome y completely
  mutate(x1 = ifelse(y == 1, x1abs, -x1abs))
# View(data_3_0)

fit_3_0 <- glm(y ~ 0 + x1 + x2 + x3,
               family = binomial("logit"),
               data = data_3_0)
```

```{r "sim_method2", eval=FALSE, include=TRUE}
n <- 100000
beta1 <- 1
beta2 <- 0.5
beta3 <- -0.25
y <- c(rep(0, n), rep(1, n))
x1 <- c(abs(rnorm(n, mean = , sd = 0.5)), 
        -abs(rnorm(n, mean = 1, sd = 0.5)))
x2 <- c(rnorm(2 * n, mean = 1, sd = 0.5))
x3 <- c(rnorm(2 * n, mean = 2, sd = 0.5))

data_3_1 <- data.frame(y, x1, x2, x3)
# mean(data_3_1$x1)
# mean(data_3_1$y)

# save(data_3_0, data_3_1, file = "bios_qexam_data_3_0.Rdata")
```


```{r "simulation", eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
n <- 100000
beta1 <- 1
beta2 <- 0.5
beta3 <- -0.25
y <- c(rep(0, n), rep(1, n))
glm_con <- matrix(NA, nrow = n, ncol = 3)
glm_unc <- matrix(NA, nrow = n, ncol = 2)

for (i in seq_along(1:1000)) {
  x1 <- c(abs(rnorm(n, mean = 1, sd = 0.5)), 
          -abs(rnorm(n, mean = 1, sd = 0.5)))
  x2 <- c(rnorm(2 * n, mean = 1, sd = 0.5))
  x3 <- c(rnorm(2 * n, mean = 2, sd = 0.5))
  
  data_sim <- data.frame(y, x1, x2, x3)
  
  fit_3_0 <- glm(y ~ 0 + x1 + x2 + x3,
               family = binomial("logit"),
               data = data_sim)
  
  fit_3_1 <- glm(y ~ 0 + x2 + x3,
               family = binomial("logit"),
               data = data_sim)
  
  glm_con[i, ] <- fit_3_0$coefficients
  glm_unc[i, ] <- fit_3_1$coefficients
}

save(glm_con, glm_unc, 
     file = "bios_qexam_data_3_0.Rdata")
```


```{r}
load("bios_qexam_data_3_0.Rdata")

library(matrixStats)
colMeans(as.matrix(glm_con), na.rm = TRUE)
colMeans(as.matrix(glm_unc), na.rm = TRUE)


colMeans(as.matrix(glm_con), na.rm = TRUE) - c(1, 0.5, -0.25)
colMeans(as.matrix(glm_unc), na.rm = TRUE) - c(1.5, -0.25)

colSds(as.matrix(glm_con), na.rm = TRUE)
colSds(as.matrix(glm_unc), na.rm = TRUE)

```


### (e). 

Perform a simulation study using the generative model in (a). 
comparing different penalization schemes. 
As in (a), we will assume a sample size of 103. 

Please use the following methods: 
(1) LASSO regression (using the glmnet package in R); 
(2) Firth correction (using the logistf package in R).

For both methods, you will actually fit the model with the intercept.
Thus, we are fitting a misspecified model, and the goal is to see how the corrections fare in terms of bias. 

For the simulation study, specify the following: 
- Tuning parameters used for the two penalized regressions; 
- The number of simulated datasets being used. 
- Report on the mean bias for the parameters using each of the methods. 

(NB: For the LASSO, you will get a warning message of the form 
"In lognet(xd, is.sparse, ix, jx, y, weights, offset, alpha, ... : one multinomial or binomial class has fewer than 8 observations; dangerous ground"; 
you can ignore that.  For the Firth correction, you will get warning messages of the form "fitted probabilities numerically 0 or 1 occurred for variable V2"; 
you can disregard those.  In the event that any of your simulation values do not converge, you may remove those from the simulation study. Finally, for the Firth correction, you will want to set the maximum number of iterations to be at least 10000).



The tuning parameter used for glmnet $\alpha = 1$ as LASSO, and $\lambda = 0.1, 0.2, and 0.5$
The tuning parameter used for logsitf $\alpha = 1$ as LASSO, and $\lambda = 0.1, 0.2, and 0.5$

```{r eval=FALSE, include=TRUE}
library(tidyverse)

sim_103 <- function(n = 103, seed){
  set.seed(seed)
  x1 <- c(abs(rnorm(n, mean = 0, sd = 0.5)))
  x2 <- c(rnorm(n, mean = 1, sd = 0.5))
  x3 <- c(rnorm(n, mean = 2, sd = 0.5))
  logit_p <- 0
  p_hat <- 1 / (1 + exp(-logit_p))
  y <- map(rep(p_hat, n), 
         ~rbinom(n = 1, size = 1, prob = .)) %>% 
  unlist()
  data_3 <- data.frame(y, x2, x3) %>%
  ## simulate x1 which can be separate by outcome y completely
  mutate(x1 = ifelse(y == 0, -x1, x1))
  
  return(data_3)
}

sim_103s <- map(1:1000, ~sim_103(seed = .))
```


```{r eval=FALSE, include=TRUE}
library(glmnet)

glmnet_01 <- map_dfc(sim_103s,
              ~glmnet(x = .[, c(4, 2, 3)],
                      y = .[, 1],
                      family = "binomial",
                      alpha = 1,
                      lambda = 0.1) %>%
                coef(s = 0.1) %>%
                as.vector())

glmnet_02 <- map_dfc(sim_103s,
              ~glmnet(x = .[, c(4, 2, 3)],
                      y = .[, 1],
                      family = "binomial",
                      alpha = 1,
                      lambda = 0.2) %>%
                coef(s = 0.2) %>%
                as.vector())


glmnet_05 <- map_dfc(sim_103s,
              ~glmnet(x = .[, c(4, 2, 3)],
                      y = .[, 1],
                      family = "binomial",
                      alpha = 1,
                      lambda = 0.5) %>%
                coef(s = 0.5) %>%
                as.vector())

save(glmnet_01, glmnet_02, glmnet_05, 
     file = "bios_qexam_question_3_glmnet.Rdata")
```

```{r}
load("bios_qexam_question_3_glmnet.Rdata")
library(matrixStats)
rowMeans(glmnet_01)
rowMeans(glmnet_02)
rowMeans(glmnet_05)
rowSds(as.matrix(glmnet_01))
rowSds(as.matrix(glmnet_02))
rowSds(as.matrix(glmnet_05))
```

$$
L^*\beta = L(\beta) \times det(I(\beta))^{1/2}
$$

where $I(\beta)$ is the Fisher information matrix and $L(\beta)$ is the likelihood.
Firth-type penalization removes the first-order bias of the MLE and bias-preventive rather than corrective for rare events. I do not exactly understand why we need to use Firth's method on noncovergence situation. There is no tuning parameter.

```{r eval=FALSE, include=TRUE}
library(logistf)
logistf_01 <- map_dfc(sim_103s,
              ~logistf(data = ., 
                   formula = y ~ 1 + x1 + x2 + x3, 
                   control = logistf.control(maxit = 1000),
                   modcontrol = logistf.mod.control(tau = 0.1),
                   firth = TRUE) %>%
                coef() %>%
                as.vector()) 

logistf_02 <- map_dfc(sim_103s,
              ~logistf(data = ., 
                   formula = y ~ 1 + x1 + x2 + x3, 
                   control = logistf.control(maxit = 1000),
                   modcontrol = logistf.mod.control(tau = 0.2),
                   firth = TRUE) %>%
                coef() %>%
                as.vector()) 

logistf_05 <- map_dfc(sim_103s,
              ~logistf(data = ., 
                   formula = y ~ 1 + x1 + x2 + x3, 
                   control = logistf.control(maxit = 1000),
                   modcontrol = logistf.mod.control(tau = 0.5),
                   firth = TRUE) %>%
                coef() %>%
                as.vector()) 

save(logistf_01, logistf_02, logistf_05, file = "bios_qexam_question_3_logistf.Rdata")
```


```{r}
load("bios_qexam_question_3_logistf.Rdata")
library(matrixStats)
rowMeans(logistf_01)
rowMeans(logistf_02)
rowMeans(logistf_05)
rowSds(as.matrix(logistf_01))
rowSds(as.matrix(logistf_02))
rowSds(as.matrix(logistf_05))

```





















